---
title: "Urban Forestry Assessment"
author: "Alexandria Reed"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    includes:
      after_body: footer.html
    code_folding: "show"
    toc: true
    toc_float: true
---

# Load Libraries

#### Various packages for wrangling, spatial analysis, and data visualization
```{r, message=FALSE}
library(tidyverse)
library(readxl)
library(sf)
library(ggplot2)
library(leaflet)
library(tmap)
library(units)
library(RColorBrewer)
```

# Read in Data

#### Working with tabular data (CSV and Excel files) and vector data (shapefiles)
```{r, results="hide"}
# Sample inventory
inventory <- read.csv("sample_data/sample_inventory.csv") 

# Sample water use ratings
rating <- read_excel("sample_data/sample_water_use_ratings.xlsx")

# Sample shapefiles for canopy, census tracts, and city boundaries
canopy <- st_read("sample_data/sample_canopy_shp/sample_canopy_polygons.shp")
census <- st_read("sample_data/sample_census_tracts/sample_census_tracts.shp")
city_bounds <- st_read("sample_data/sample_city_boundary/sample_city_boundary.shp")
```

# Data Exploration

#### Familiarizing myself with the tabular data and using `tmap` to view the shapefiles
```{r, results="hide", message=FALSE}
# Exploring tabular data
summary(inventory)
# Platanus_x_acerifolia and Platanus_x_hispanica have the same common name

summary(rating)

# Viewing vector data 
tmap_mode("view") +
tm_shape(canopy) +
  tm_polygons()

tmap_mode("view") +
tm_shape(census) +
  tm_polygons()

tmap_mode("view") +
tm_shape(city_bounds) +
  tm_polygons()
```

# Visualization that shows the percent of trees in the inventory that fall into each water use rating category based on species.

#### Calculating percent of trees in each water use rating category based on species
```{r, message = FALSE}
# Calculate the percent of each water use rating category for each species
rating_percent <- inventory |>
  left_join(rating, by = c("scientific_name" = "species")) |>
  group_by(scientific_name, water_use_rating) |>
  summarize(percent = n() / nrow(inventory) * 100)
```


#### Creating a bar plot using `ggplot2`
```{r, message=FALSE, fig.align="center", fig.width=8, fig.height=6, fig.cap="Figure 1: This visualization illustrates the proportion of trees within the inventory categorized by their water use rating, segmented by species."}
# Wrangling for plot
# Change special value to character string so that I can change the order
rating_percent[6, 2] <- "NA"

# Reorder the levels of water_use_rating based on custom order
custom_order <- c("NA", "High", "Low", "Moderate", "Very Low")
rating_percent$water_use_rating <- fct_relevel(rating_percent$water_use_rating, custom_order)

# Create the plot
ggplot(rating_percent, aes(x = water_use_rating, y = percent, fill = scientific_name)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(percent, "%"), group = scientific_name), 
            position = position_stack(vjust = 0.5), size = 3.5, color = "black") +
  labs(title = "Distribution of Trees by Water Use Rating and Species",
       x = "Water Use Rating", 
       y = "Trees (%)") +
  theme_minimal() +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  guides(fill = guide_legend(title = "Species")) +
  scale_fill_brewer(palette = "BrBG") 
```

```{r, include=FALSE}
# Not included in html
# One approach, but doesn't answer the prompt as clearly

# Calculate the percentage of each water use rating category for each species
rating_percent <- inventory |>
  left_join(rating, by = c("scientific_name" = "species")) |>
  group_by(scientific_name, water_use_rating) |>
  summarize(percent = n() / nrow(inventory) * 100)

# Sort percent in descending order
rating_percent <- rating_percent[order(rating_percent$percent, decreasing = TRUE), ]

# Change special value to character string so that I can give it a specific color
rating_percent[3, 2] <- "NA"

# Create the plot
ggplot(rating_percent, aes(x = reorder(scientific_name, percent), y = percent, fill = water_use_rating)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("High" = brewer.pal(9, "Blues")[7],
                               "Moderate" = brewer.pal(9, "Blues")[6],
                               "Low" = brewer.pal(9, "Blues")[5],
                               "Very Low" = brewer.pal(9, "Blues")[4],
                               "NA" = "grey"),
                                        limits = c("High", "Moderate", "Low", "Very Low", "NA")) +
    labs(title = "Distribution of Trees by Water Use Rating and Species",
       x = "Species", 
       y = "Trees (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  geom_text(aes(label = paste0(percent,"%"), y = percent), 
            hjust = -0.15,  
            size = 3.5) +  
  guides(fill = guide_legend(title = "Water Use Rating")) +
  ylim(0, 33)  
```

# Map showing the canopy cover percent for each council district.

#### Geospatial analysis using `sf`, `dplyr`, and `units`
```{r, warning=FALSE}
# Calculate the intersection of census tracts and canopy polygons
intersection <- st_intersection(census, canopy)

# Calculate canopy area (convert m² to acres)
intersection$canopy_area_acres <- st_area(intersection) / 4046.86

# Calculate canopy cover for each census tract
canopy_cover <- intersection |>
  group_by(tract) |>
  summarise(canopy_area = sum(canopy_area_acres)) |>
  st_drop_geometry() # drop so that we can merge seamlessly

# Merge the canopy cover data with the census data 
census_w_canopy_cover <- merge(census, canopy_cover, by = "tract")

# Drop units for clarity
census_w_canopy_cover <- drop_units(census_w_canopy_cover)

# Calculate the percentage of canopy cover for each census tract
census_w_canopy_cover$canopy_cover_percent <- (census_w_canopy_cover$canopy_area / census_w_canopy_cover$acres) * 100
```

#### Using `leaflet` to create the map

<style>
.leaflet {
    margin: auto;
}
</style>

```{r}
# Create the leaflet map
color_pal <- colorNumeric(palette = "Greens", domain = census_w_canopy_cover$canopy_cover_percent)

leaflet(census_w_canopy_cover) |>
  addProviderTiles("Esri.WorldTopoMap") |>
  addPolygons(fillColor = ~color_pal(canopy_cover_percent),
              fillOpacity = 0.7,
              color = "black",
              weight = 1,
              popup = paste("Tract ID: ", census_w_canopy_cover$tract, "<br>",
                            "Canopy Cover: ", paste0(round(census_w_canopy_cover$canopy_cover_percent, 2), "%"))) |>
  addLegend(pal = color_pal,
            values = ~canopy_cover_percent,
            title = "Canopy Cover (%)",
            position = "bottomright")
```

# Function for Calculating Total Canopy Area

#### Creating the function that calculates the total canopy area (in square feet) of the tree inventory based on the canopy width attribute
```{r}
# Formula to calculate the area of a circular canopy is pi * (radius^2)
# Function to calculate total canopy area
calculate_total_canopy_area <- function(inventory) {
  total_area <- sum(pi * (0.5 * inventory$canopy_width_ft)^2)
  return(total_area)
}

# Calculate total canopy area
total_canopy_area <- calculate_total_canopy_area(inventory)

# Print the total canopy area
cat("The total canopy area is", format(total_canopy_area, big.mark = ",", scientific = FALSE), "square feet.")
```

